---
title: 'Session 1: Organization, Introduction to R, Data Import'
output: html_document
date: "2023-01-11"
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Organization

## Ressources

- R Markdown notebooks are located here <https://github.com/gmoeser> - please download first.
- There are 6 R Markdown notebooks, one for each session. This is the first one.
- Dataset is located here: <https://raw.githubusercontent.com/gmoeser/R-lectures/master/ElectricCarData_Clean.csv>
   
   
## References: 
- R Graphics Cookbook: <https://r-graphics.org/recipe-quick-line>
- R Practice: https://bookdown.org/logan_kelly/r_practice/p09.html
   
   
## Some notes about R Markdown notebooks

R Markdown notebooks are organized along two kinds of sections:

- Markdown sections (like this one) and
- Code chunks (or sections) which support several programming languages, like R, Python, SQL etc.
- Notebooks can be knitted into MS Word, HTML or other supported formats. Just press the blue knit-button.
   
   
## Code chunks
   
...start with three back ticks and the programming language in curly brackets, see the example. Code chunks can be run with the *green play-button* on the left.
   
```{r ExampleRCodeChunk}

```
   
   
## Exercise: R, RStudio, RMarkdown

- [if not already done] Please download and install R, latest version
- [if not already done] Please download and install RStudio, latest version
- Please note: both software package are available as open source!
- Please download the two RMarkdown documents 
- Open both RMarkdown documents in RStudio
- Try to *knit* the first RMarkdown document (Session1.Rmd) into a html-document
- Add a code chunk in the document 
   
   
**Please note:** ipads are currently not supported, if running an ipad (or an Android pad), please go to https://posit.cloud. Currently, I cannot offer any support here!
   


# Introduction to R

## R is a vector oriented - statistical - programming language

R is a vector oriented language: all operations will be performed on the
whole vector at once. It is not necessary to loop over each element in a
vector and perform - if a certain condition is fullfilled - a certain
action.

```{r IntroductionToR}
# Generate a small vector

# Perform an operation on the vector: multiply each element by 2

```

You can see that each element in vector `MyFirstVectorInR` was
multiplied by value 2. No need for any loops or `if`-clauses. You could
also see how a vector was set up: use the `c()` function and assign `<-`
it to an object, here like `MyFirstVectorInR`. Normally, it is not
necessary to set up vectors manually. Data will be imported using a
function like `read.table()`.

## Everything is an object!

In R, we can store everything into an object using the assignment
operator: `<-` To access an object, just enter its name on the console.

```{r}

# Access the object

```

## Help

R has a built-in help system.

```{r}
# usage of the built-in help system

# faster

```

Another very important ressource is Stack Overflow:
<https://stackoverflow.com/questions/tagged/r>

In RStudio, in the lower right pane, a **Help tab** is available. Here,
a search function is available.

## Packages

A lot of additional R packages are available.

```{r}
# Currently installed packages

```

Install a package either using the `install.packages()` function or use
the RStudio interface, which can be found in tab `Packages` in the
lower-right pane in RStudio.

```{r}
# Install a package: lm.beta

```

After a package was successfully installed, the whole package can be
loaded using `library()` function or single functions can be used using
the package name followed by two colons -
`packageName::functionName()` - followed by the functions name, see
example below.

```{r}
# Load thw complete package

```

Use only a single function from a package:

```{r}
# Use only a single function to get the standardized regression coefficient from a simple linear regression

```

## Exercises: Introduction to R

1.  What is the function to calculate the standard deviation? Play
    around with the examples of the help file.
2.  How to generate a boxplot? Play around with the examples of the help
    file.
3.  What is `CRAN`?
4.  How many official packages are currently available on CRAN?
5.  Please install the `psych` and `Hmisc` packages. Try to get an
    overview of both packages. See, if there is are so called
    `vignettes` available. Try to find also out what a `Vignette` is.
6.  Load the `psych` package
7.  Use the `describe()` function in `Hmisc` package to get some
    information about the `mtcars` dataset. First, open the help page of
    the `mtcars` dataset. It is not necessary to import the `mtcars`
    dataset, it is already available. Why?

Please note: Add more chunks here if needed!

```{r}

```
    
    
# Importing Data

The dataset is stored here:
<https://raw.githubusercontent.com/gmoeser/R-lectures/master/ElectricCarData_Clean.csv>

Let us put the URL in an object first, using the assignment operator
`<-`:

```{r URL}
# Assign URL of data into object URL
URL <- "https://raw.githubusercontent.com/gmoeser/R-lectures/master/ElectricCarData_Clean.csv"
```

Import the data into global environment:

```{r ImportDataset}

```
   
You can get a first impression of the importet object using the `head()` or `tail()` functions (and in RStudio `View`):  
  
```{r}
# head will print out the first six cases

```

     
## Exercises: Importing Data

- Please import the diamonds dataset from
  <https://raw.githubusercontent.com/tidyverse/ggplot2/main/data-raw/diamonds.csv>
- Store the dataset in an object called `diamonds`
   
   
# Data Structures in R

R has essentially only two data structures:

1.  vector
2.  list

But we want to work with a **statistical table**, which contains data
organized in a specific way:

- In rows are units, like the individual car type
- In columns are variables, like the average range of a car, price of a car etc.
- In cells are values

This structure is available in R as a so called `data.frame`. A `data.frame` is a `list` organized in the - described above - specific data structure. The imported data is stored directly into a `data.frame`:

```{r DataStrcuturesInR}
# We use indexing to print out only a few cases and variables

```

You can see that the data is structured in the above described ways. In
first row is a Tesla stored, which you can easily find out checking the
variables `Brand` and `Model` in first and second column. The values
stored in cells are then `Tesla` and `Model 3 Long Range Dual Motor`.
You will find more related information stored in other variables like
`Range_Km` etc.

## Access a single vector (variable)

A single variable in a `data.frame` is an object type called `vector`. A
vector can be assigned in different ways, one qay is to use the `$`
sign: first, enter the name of the dataset, followed by the `$` sign,
followed by the name of the variable: `NameOfDataFrame$VariableName`

```{r}
# We want to assign the AccelSec variable and calculate the arithmetic mean

```

There are more ways how to access a single variable in R, here are some
examples (will not discuss them in details now):

```{r}
# Two more ways to select variables from a list or data.frame

# ...
```
   
   
   
## Exercises: Data Structures

- Try to find out, if the imported object `diamonds` (last exercise) is a data.frame? (Hint: Use `str()` or `is.data.frame()` or ...)
- Access a single vector in the `diamonds` objects, e.g. carat, and calculate the average mean and standard deviation


Please note: Add more chunks here if needed!

```{r}

```

