---
title: 'Session 2: Data Understanding'
output: html_document
date: "2023-01-11"
editor_options: 
  markdown: 
    wrap: 72
always_allow_html: true    
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# Required packages
#install.packages("plotly")

```

# Data Understanding: Descriptive Analyses and Charts

## Data Quality

Before performing hypothesis testing, multivariate models etc. it is important to check the **data quality**:   

- Are there any outliers? 
- Are there any unplausible values? 
- Are there missing values? 
- Of what kind is the distribution of the single variables? 
- ... 
   


## Built-in functions: `summary()` and `str()`

R provides two very helpful built-in functions: `str()` and `summary`,
which should be applied to a new dataset directly after importing it.

Import the dataset (please note: new RMarkdown document --\> need to
import the dataset again)

```{r URL}
# Assign URL of data into object URL
URL <- "https://raw.githubusercontent.com/gmoeser/R-lectures/master/ElectricCarData_Clean.csv"
# and import the dataset

```

First check of the data using the `str()` function.

```{r}

```

Here, you can see that the `ElectricVehicleData` object is a
`data.frame` and that we have 103 observations (rows) and 14 variables
(columns). You can see the name of the variables and the storage types
(like `chr`, `num` or `int`). Also, a few values (cells) are shown.

Next, we can request some basic statistical paramater applying the
`summary()` function:

```{r}

```

The `summary()` function provides - based on the storage type - some
basic statistical parameters.

### A note about storage types and measurement levels

Storage types (like num, int, char etc.) and not directly related to
measurement levels (Stevens, 1946):     
   
- `num` and `int` could be related to measurement levels `ratio` and `interval` 
- If an `int` variable
represents a Likert-scale, then it is related to `ordinal` measurement
level 
- `char` is typically related to `nominal` scaled variables (but
an `int` variable can be used as well) 
- There are two special types,
called `factor` and `ordered`. `ordered` represents an ordered
categorical variable, an `ordinal` scaled variable, while `factor`
typically represents an unordered categorical variable, a `nominal`
scaled variable.
    
    
Especially with **ordinal scaled** variables, it is often necessary to
perform some transformations before an arithmetic mean can be
calculated. Please note that an arithmetic mean shouldn't be calculated
for a ordinal scaled variable, but nevertheless, this is what happens in
reality...

   
   
### Exercises: Built-in functions, Missing Values

- Please import the diamonds dataset, https://raw.githubusercontent.com/tidyverse/ggplot2/main/data-raw/diamonds.csv
- Please inspect the dataset using:
  - `str()` - which parameters are shown?
  - `summary()`- which parameters are printed out depending on what?
  - RStudio (upper right pane, look for the object) and `View()` 
  - `head()` - change the number of rows to 10
  - `tail()`
- Apply the `summary()` function on the `airquality` dataset. Check the help page for the `airquality` first. What is special about the variables `Ozone` and `Solar.R`?


```{r}

```
   

  

## Charts

There are several ways how to get visualizations in R:   
   
   
- Built-in charts 
- Additional packages, like ggplot2 
- Interactive charts with plotly
- ...
  


### Built-in charts

Built-in charts can be used during the data analysis process to get
insights about the data.
   
   
#### Barplot

A barplot in R is based on a frequency table of a variable. A frequency
table can be generated using `table()` function.

```{r}
# Distribution of PlugType



```
   
   
#### Boxplot

A boxplot can be drawn for measurement levels `ordinal`, `interval` and
`ratio`.

```{r}

```
    
    
#### Histogram and Density Plots

A histogram can be drawn for measurement levels `interval` and `ratio`

```{r}

```

Together with a density plot

```{r}


```
   
       
#### Multiple plots side-by-side  
  
Two boxplots side-by-side. Please note that this works in the output
html-document, it will displayed here slightly different.

```{r, figures-side, fig.show="hold", out.width="50%"}


```

### Interactive charts can be requested with `plotly`.

We can request interactive charts as well, using `plotly` for example.

Interactive scatterplot with plotly:

```{r}


```
   
   
   
### Exercises: Charts
    
    
- Generate a barplot, a boxplot and a histogram (with density plot overlayed) for a variable of your choice in the electric vehicles data.frame.   

- Generate an interactive chart with plotly between two variables of your choice in electric vehicles dataset or in diamonds dataset.
   
   
```{r}

```

   
   
## Descriptives

It is simple to generate classical descriptive parameters in R for
variables. Typical functions are `min()`, `max()`, `median()`, `mean()`,
`sd()` etc.

Simple example:

```{r}

```

But how to get more statistical parameter for more variables at once?

First, we can use a named vector to put the different statistical
parameters in.

```{r}




```

Second, we can put this into a function in R to apply it to any variable

```{r}





```

Third, we can use a loop to apply out function to a set of variables

```{r}

```

Of course, wen can store the result into a data.frame, transpose it (if
you want to switch rows and columns) and export it in MS Excel format.
   
   
## Exercise: Data Understanding: Descriptive Analyses

- Calculate the arithmetic mean, the standard deviation and minimum/maximum for `Ozone` and `Solar.R` (`airqualiy` data.frame). What is the output and why? How to get a numerical result? Check the help page for more information.
- Set up your own function with parameters Min / Max / Median / Mean and Standard Deviation
- Apply the function to a few variables in the diamonds dataset   
   
```{r}

```


